Evolution of ORM Systems: Formal Foundations

### 2013-01-21 doc/description.tex
small fixes

The paper is organized as follows: in Section \ref{sec:evoIntro} the concept of ORM software and its evolution is discussed, then models of application and database are introduced in Section \ref{sec:appModel} and in Section \ref{sec:dbModel}, the ORM is defined in Section \ref{sec:orm} and the evolutionary transformations are defined in Section \ref{sec:eorm}.

@@

The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user. The components of a software (application and database) are transformed likewise, thus the state is defined for them too.
\begin{align*}
 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s,  \\	s \lhd Software \vee s \lh Application \vee s \lhd Database
\end{align*}
Before the definition of the software consistency is provided, the notation of function transcription has to be introduced. Usually the definition of a function consists of two parts - an effect of the function and conditions when the effect occurs We decide to use following transcription, to simplify the notation and to improve understanding of functions:
\begin{align*}
f(x) =   \inference{conditions}{effect}
\end{align*}

The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
 	\mathbf{consistent} : State \rightarrow Boolean \\
 	consistent(s) = \begin{cases}
 		\inference{s = Consistent(s)}{true}\\\\
 		\inference{s = Inconsistent(s)}{false}\\
 \end{cases}\\
 s \lhd State
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
	\mathbf{state} : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ 
 	s \lhd Software
\end{align*}

@@

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
$$\oplus : State \times E \rightarrow State $$
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
\circ : State \times AppT \rightarrow State \\
 \bullet : State \times DbT \rightarrow State
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.
\begin{align*}
 state(s) \oplus t = \begin{cases}
 	\inference{! consistent(state(s))}{u} \\ \\
 	\inference{consistent(state(s))}{state(sw(\Psi(s, t), \Phi(s,t))} 
\end{cases}\\ 
s \lhd Software, t \in E
\end{align*}
\begin{align*}
 state(a) \circ e = \begin{cases}
 	\inference{!\;consistent(state(a))}{a} \\ \\
 	\inference{consistent(state(a))}{e(a)} 
\end{cases}\\ 
a \lhd Application, e \in AppT
\end{align*}
\begin{align*}
 state(d) \bullet e = \begin{cases}
 	\inference{!\;consistent(state(d))}{d} \\ \\
 	\inference{consistent(state(d))}{e(d)} 
\end{cases}\\ 
d \lhd Application, e \in DbT
\end{align*}

@@

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pkValue, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, n, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, n, ps)) = ps \\ \\
	\mathbf{pkValue} : Row \rightarrow N \\
	pkValue(row(l, n, ps)) = n \\ \\
	 l \in Label, ps = Pair*, n \in N
\end{align*}
The value stored in concrete column of the row is represented by a pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Value \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		%\inference{r = \varnothing}{\varnothing} \\
 		%\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \lhd Row*, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}

@@

\paragraph{Add Value To Table} The function appends a pair to all rows of a table.
\begin{align*}
	\mathbf{addValueToTable} : Label \times Pair \times Database \rightarrow Database \\
	addValueToTable(l, p, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l}	{\begin{gathered}
		d \bullet insertValue(head(tableData(l, d)), p, d) \\ \bullet addValueToTable(tail(tableData(l, d)), p, d)
	\end{gathered}} \\ \\
 		Inconsistent \; d
 	\end{cases}\\ \\
 	l \in Label, p \lhd Pair,  d \lhd Database
\end{align*}

@@

\subsubsection{Database decomposition}
\paragraph{dropValue}
\begin{align*}
	\mathbf{dropValue} : Label \times Row \times Database \rightarrow Database \\
	dropValue(l, r, d) = \inference{\exists p \in pairs(r) : reference(p) = l}{pairs(r) \setminus p}
	\\ l \in Label, r \lhd Row \wedge r \in data(d), d \lhd Database
\end{align*}

\begin{align*}
	\mathbf{dropRow} :  Row*  \times Database \rightarrow Database\\
	dropRow(r, d) = \inference{|r|=1}{data(d) \setminus r}
	\\ l \in Label, r \lhd Row \wedge r \in data(d), d \lhd Database
\end{align*}

\paragraph{Remove table} The function removes a table if there are no stored data. This is because the data lost is undesirable in case of evolution. The consistency of foreign keys references has to be verified too.
\begin{align*}
	\mathbf{dropTable} : Label \times Database \rightarrow Database \\
	dropTable(l,d) = \inference{ \exists t \in tables(d) : name(t) = l  \\ \wedge \forall r \in data(d) : owningTable(r) \neq l \\  \wedge (\forall j \in tables(d) \wedge f \in foreignKeys(j) : reference(f) \neq l) }{ tables(d) \setminus \rho(c)}
\end{align*}

\begin{align*}
	\mathbf{dropTableAndData} : Labe \times Database \rightarrow Database \\
	dropTableAndData(l, d) = \inference{\exists t \in tables(d) \\ \wedge ! \; associated(t, d)}{\begin{gathered}
	(\forall r \in tableData(t,d) : dropRow(r, d))\\ \bullet dropTable(t,d)
\end{gathered}}
	\\ l \in Label, d \lhd Database
\end{align*}

\paragraph{Remove column or foreign key} The situation is similar to the $dropTable$ function - there has to be no data in the column or key so it can be safely removed.
\begin{align*}
	\mathbf{dropColumn} : Label \times Label \times Database \rightarrow Database \\
	dropColumn(l, l_p, d) = \inference{ \exists t \in tables(d) : name(t) = l \\ \wedge \exists c \in columns(t) : name(c) = l_p \\ \wedge \forall e \in  pairs(r), r \in tableData(t, d) : \\ reference(e) \neq l_p \vee value(e) = \varnothing  }{columns(t) \setminus c}\\
	 l , l_p\in Label, d \lhd Database 
\end{align*}

\begin{align*}
	\mathbf{dropKey} : Label \times Label \times Database \rightarrow Database \\
	dropKey(l, l_p, d) = \inference{ \exists t \in tables(d) : name(t) = l \\ \wedge \exists c \in foreignKeys(t) : name(c) = l_p \\ \wedge \forall e \in  pairs(r), r \in tableData(t, d) : \\ reference(e) \neq l_p \vee value(e) = \varnothing  }{foreignKeys(t) \setminus c}\\
	 l, l_p \in Label, d \lhd Database 
\end{align*}

\paragraph{Remove Data and Column} The approach we declare to introduce in this paper was declared to be data and information preserving, nevertheless there are cases when data can be deleted from the database without data lost as they are e.g. already copied in another column as the understanding of data changed in the whole software. 
\begin{align*}
	\mathbf{DropColumnAndData} : Label \times Label* \times Database \rightarrow Database \\
	DropColumnAndData(l_c, ps, d) = \begin{cases}
 	%\inference{ps = \varnothing}{d}\\
 	 %\inference{|ps| >1}{\begin{gathered} 	d \bullet DropColumnAndData(l_c, head(ps), d) \\ \bullet DropColumnAndData(l_c, tail(ps), d) \end{gathered}} \\ \\
 	(\inference{|ps| = 1 \\ \wedge \exists t \in tables(d) \\ \wedge \exists c \in columns(t) : name(c) = l_c}{\begin{gathered}
	 	(\forall r, r \in tableData(d) \wedge reference(r) = l_c: \\d \bullet dropValue(ps, r)) \end{gathered}}) \\ \bullet dropColumn(l_c)  
 \end{cases}
	\\
	l_c \in Label, ps \lhd Label*, d \in Database
\end{align*}

@@

\section{Evolutionaty Transformations}
The set of all possible data evolution cases $E$ was defined in Section \ref{sec:evoIntro}. The semantic of transformations in the set $E$ is defined by their interpretation on individual software components. This section provides an overview of basic transformations (evolution cases) and theirs interpretation on the application and the database. The section starts with simple transformations as the complex one can be often implemented as theirs concatenation. The transformations are divided into three groups - first group creates the structure of a software, second group removes parts of a software structure and third group is able to alter the structure.

@@

\subsection{Software Creation}
There are defined three transformations which are able to create a software: $addProperty$, $addAssociation$,  $newClass$. These three transformations can be used to create a software from a scratch. The last of them has no impact on stored data, whereas the rest of them can in some cases.
\paragraph{Add Property} Adding a property to database can change not only structure but stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{align*}
	\mathbf{addProperty} : Label \times Property \times State \rightarrow State \\
	\Psi(s, addProperty(l, p)) = application(s) \circ addProp(l,p) \\
	\Phi(s, addProperty(l, p)) = \begin{cases}
    		\inference{cardinality(p) \leq 1}{database(s) \bullet addColumn(l, \rho(p,d), d)} \\ \\
	    \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
   database(s) \bullet addTable(\rho(p,d), d) \\ \bullet \forall r \in row(owningClass(p)) :\\ insertRow(row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))])) 
    \end{gathered} } 
   \end{cases}\\
	s \lhd Software, l \in Label, p \in Properties
\end{align*}

\paragraph{Add association} Adding an association into a database is complicated because of already existing instances. When an association between classes is added the instances of both classes has to be connected on the database level via foreign key or connection table. The information about pairing has to be provided by the user in form of set of instances pairs, because the model does not contain enough information. 
\begin{align*}
\mathbf{addAsociation} : Label \times Association \times (Row \times Row) \times State \rightarrow State \\
\Psi(s, addAsociation(l, a, m_{l, reference(a)})) = application(s) \circ addAssoc(l, a) \\
\Phi(s, addAsociation(l, a, m_{l, reference(a)})) = \begin{cases}
	\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = l\\ 		\wedge name(t_t) = reference(a) \\ \wedge \forall f \in foreignKeys(t_t) \cup columns(t_t) \\: name(f) \neq name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
		database(s) \\ \bullet addForeignKey(name(t_s), \rho(a,d), d) \\ \bullet \forall r \in m_{l, reference(a)} : \\ insertRow(row(name(t_s), [name(a), \\ value(primaryKey(first(r)))]), d)
	\end{gathered}}
 \\ \\
	\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = l \\ \wedge name(t_t) = reference(a) \\ \wedge startCardinality(a) > 1}{\begin{gathered} 
	database(s) \bullet addTable(\rho(a,d),d) \\ \bullet \forall r \in m_{l, reference(a)} :\\ insertRow(row(name(owningClass(a)) + name(a), \\ [name(t_s), value(primaryKey(first(r)))]\\ [name(t_t), value(primaryKey(second(r)))]))
	\end{gathered}} 
\end{cases}\\
\end{align*}

@@

\paragraph{Remove property}
\begin{align*}
	\mathbf{RemoveProperty} : Label \times Label \times State \rightarrow State \\
	\Psi RemoveProperty(l, l_p, s) =  applications(s) \circ  remProp(l, l_p) \\
	\Phi RemoveProperty(l, l_p, s) = \begin{cases}
		 \inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) \leq 1}{database(s) \bullet dropColumn(l, l_p)}  \\\\
		 \inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists p \in properties(c)  : name(p) = l \\ \wedge cardinality(p)>1}{database(s) \bullet dropTable(name(owningClass(p)+l)}
 	\end{cases}\\
 	l, l_p \in Label, s \lhd Software
\end{align*}

\paragraph{Remove Property and its Data} The $RemovePropertyAndData$ transformation helps to remove a property together with data stored in the database. The evolution can cause data and information lost, to prevent this the transformation $RemovePropertyAndData$ should be used carefully.
\begin{align*}
	\mathbf{RemovePropertyAndData} : Label \times Label* \times State \rightarrow State \\
	\Phi(RemovePropertyAndData(L_c, ps), s) = application(s)\\ \circ remProperty(l_s, ps) \\
	\Psi(RemovePropertyAndData(L_c, ps), s) = \begin{cases}
		%\inference{ps = \varnothing}{d}\\
		%\inference{|ps| > 1}{\begin{gathered} database(s) \\ \bullet RemovePropertyAndData(L_c, head(ps)) \\ \bullet RemovePropertyAndData(L_c, tail(ps)) 	\end{gathered}}\\\\
 		\inference{\exists c \in classes(application(s)) : name(c) = l\\ 	\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) \leq 1}{database(s) \bullet DropColumnAndData(l_c, ps)}\\\\
 	\inference{\exists c \in classes(application(s)) : name(c) = l\\ 	\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) > 1}{database(s) \bullet RemoveTableAndData(l_c, ps)}
 \end{cases}
	\\
	l_c \in Label, ps \lhd Label*, s \lhd Software
\end{align*}

\paragraph{Remove association} 
\begin{align*}
    \mathbf{RemoveAssociation} : Label \times Labels \times State \rightarrow State \\ 
    \Psi RemoveAssociation(l, l_a, s) = application(s) \circ remAss(l, l_a) \\
    \Phi RemoveAssociation(l, l_a, s) = \begin{cases}
     	\inference{\exists c \in classes(application(s)) : name(c) = l \\
     	\wedge \exists a \in associaions(c) : name(a) = l_a  \\
     	\wedge startCardinality(a) \leq 1 \\}{database(s) \bullet dropForeignKey(l_a, reference(a))} \\\\
     	\inference{\exists c \in classes(application(s)) : name(c) = l \\
     	\wedge \exists a \in associaions(c) : name(a) = l_a  \\
     	\wedge startCardinality(a) > 1}{database(s) \bullet dropTable(name(owningClass(a))+l_a)} 
     \end{cases}\\
    l_c, l_a \in Labels, s \lhd Software
\end{align*}

>>>>

The paper is organized as follows: in Section \ref{sec:evoIntro} the concept of ORM software and its evolution is discussed, then models of application and database are introduced in Section \ref{sec:appModel} and in Section \ref{sec:dbModel}, the ORM is defined in Section \ref{sec:orm} and the evolutionary transformations are defined in Section \ref{sec:Evolutionaty-Transformations}.

@@

The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user. The components of a software (application and database) are transformed likewise, thus the state is defined for them too.
\begin{align*}
 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s,  \\	s \lhd State \vee s \lh Application \vee s \lhd Database
\end{align*}
Before the definition of the software consistency is provided, the notation of function transcription has to be introduced. Usually the definition of a function consists of two parts - an effect of the function and conditions when the effect occurs We decide to use following transcription, to simplify the notation and to improve understanding of functions:
\begin{align*}
f(x) =   \inference{conditions}{effect}
\end{align*}

The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
 	\mathbf{consistent} : State \rightarrow Boolean \\
 	consistent(s) = \begin{cases}
 		\inference{s = Consistent(s)}{true}\\\\
 		\inference{s = Inconsistent(s)}{false}\\
 \end{cases}\\
 s \lhd State
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
	\mathbf{state} : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ 
 	s \lhd State
\end{align*}

@@

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
$$\oplus : State \times E \rightarrow State $$
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
\circ : State \times AppT \rightarrow State \\
 \bullet : State \times DbT \rightarrow State
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.
\begin{align*}
 state(s) \oplus t = \begin{cases}
 	\inference{! consistent(state(s))}{u} \\ \\
 	\inference{consistent(state(s))}{state(sw(\Psi(s, t), \Phi(s,t))} 
\end{cases}\\ 
s \lhd State, t \in E
\end{align*}
\begin{align*}
 state(a) \circ e = \begin{cases}
 	\inference{!\;consistent(state(a))}{a} \\ \\
 	\inference{consistent(state(a))}{e(a)} 
\end{cases}\\ 
a \lhd Application, e \in AppT
\end{align*}
\begin{align*}
 state(d) \bullet e = \begin{cases}
 	\inference{!\;consistent(state(d))}{d} \\ \\
 	\inference{consistent(state(d))}{e(d)} 
\end{cases}\\ 
d \lhd Application, e \in DbT
\end{align*}

@@

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pkValue, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, n, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, n, ps)) = ps \\ \\
	\mathbf{pkValue} : Row \rightarrow N \\
	pkValue(row(l, n, ps)) = n \\ \\
	 l \in Label, ps = Pair*, n \in N
\end{align*}
The value stored in concrete column of the row is represented by a pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Value \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		%\inference{r = \varnothing}{\varnothing} \\
 		%\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \lhd Row*, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd State 
\end{align*}

@@

\paragraph{Add Value To Table} The function appends a pair to all rows of a table.
\begin{align*}
	\mathbf{addValueToTable} : Label \times Pair \times Database \rightarrow Database \\
	addValueToTable(l, p, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l}	{\begin{gathered}
		d \bullet insertValue(head(tableData(l, d)), p, d) \\ \bullet addValueToTable(tail(tableData(l, d)), p, d)
	\end{gathered}} 
 	\end{cases}\\ 
 	l \in Label, p \lhd Pair,  d \lhd Database
\end{align*}

@@

\subsubsection{Database decomposition}
\paragraph{Remove a value}  The approach introduced in this paper has to  preserve data and information, nevertheless there are cases when data can be deleted from the database without data lost as they are e.g. already copied in another column as the understanding of data changed in the whole software, therefore we defined two transformations - one removes a single value from a row, second removes whole row (instance) from the software.
\begin{align*}
	\mathbf{dropValue} : Label \times Row \times Database \rightarrow Database \\
	dropValue(l, r, d) = \inference{\exists p \in pairs(r) : reference(p) = l}{pairs(r) \setminus p}
	\\ l \in Label, r \lhd Row \wedge r \in data(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{dropRow} :  Row*  \times Database \rightarrow Database\\
	dropRow(r, d) = \inference{|r|=1}{data(d) \setminus r}
	\\ l \in Label, r \lhd Row \wedge r \in data(d), d \lhd Database
\end{align*}

\paragraph{Remove table, column or key} The functions removes a table, column or key if there are no stored data. This is because the data lost is undesirable in case of evolution. The consistency of foreign keys references has to be verified too in case a table is removed.  There are defined two variants of remove transformations for each model entity - one preserves data, the other removes data too.

\begin{align*}
	\mathbf{dropTable} : Label \times Database \rightarrow Database \\
	dropTable(l,d) = \inference{ \exists t \in tables(d) : name(t) = l  \\ \wedge \forall r \in data(d) : owningTable(r) \neq l \\  \wedge (\forall j \in tables(d) \wedge f \in foreignKeys(j) : reference(f) \neq l) }{ tables(d) \setminus \rho(c)}
\end{align*}

\begin{align*}
	\mathbf{dropTableAndData} : Labe \times Database \rightarrow Database \\
	dropTableAndData(l, d) = \inference{\exists t \in tables(d) \\ \wedge ! \; associated(t, d)}{\begin{gathered}
	(\forall r \in tableData(t,d) : dropRow(r, d))\\ \bullet dropTable(t,d)
\end{gathered}}
	\\ l \in Label, d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{dropColumn} : Label \times Label \times Database \rightarrow Database \\
	dropColumn(l, l_p, d) = \inference{ \exists t \in tables(d) : name(t) = l \\ \wedge \exists c \in columns(t) : name(c) = l_p \\ \wedge \forall e \in  pairs(r), r \in tableData(t, d) : \\ reference(e) \neq l_p \vee value(e) = \varnothing  }{columns(t) \setminus c}\\
	 l , l_p\in Label, d \lhd Database 
\end{align*}

\begin{align*}
	\mathbf{dropKey} : Label \times Label \times Database \rightarrow Database \\
	dropKey(l, l_p, d) = \inference{ \exists t \in tables(d) : name(t) = l \\ \wedge \exists c \in foreignKeys(t) : name(c) = l_p \\ \wedge \forall e \in  pairs(r), r \in tableData(t, d) : \\ reference(e) \neq l_p \vee value(e) = \varnothing  }{foreignKeys(t) \setminus c}\\
	 l, l_p \in Label, d \lhd Database 
\end{align*}
\begin{align*}
	\mathbf{DropColumnAndData} : Label \times Label* \times Database \rightarrow Database \\
	DropColumnAndData(l_c, ps, d) = \begin{cases}
 	%\inference{ps = \varnothing}{d}\\
 	 %\inference{|ps| >1}{\begin{gathered} 	d \bullet DropColumnAndData(l_c, head(ps), d) \\ \bullet DropColumnAndData(l_c, tail(ps), d) \end{gathered}} \\ \\
 	(\inference{|ps| = 1 \\ \wedge \exists t \in tables(d) \\ \wedge \exists c \in columns(t) : name(c) = l_c}{\begin{gathered}
	 	(\forall r, r \in tableData(d) \wedge reference(r) = l_c: \\d \bullet dropValue(ps, r)) \end{gathered}}) \\ \bullet dropColumn(l_c)  
 \end{cases}
	\\
	l_c \in Label, ps \lhd Label*, d \in Database
\end{align*}

@@

\section{Evolutionaty Transformations}
\label{sec:Evolutionaty-Transformations}
The set of all possible data evolution cases $E$ was defined in Section \ref{sec:evoIntro}. The semantic of transformations in the set $E$ is defined by their interpretation on individual software components. This section provides an overview of basic transformations (evolution cases) and theirs interpretation on the application and the database. The section starts with simple transformations as the complex one can be often implemented as theirs concatenation. The transformations are divided into three groups - first group creates the structure of a software, second group removes parts of a software structure and third group is able to alter the structure.

@@

\subsection{Software Creation}
There are defined three transformations which are able to create a software: $addProperty$, $addAssociation$,  $newClass$. These three transformations can be used to create a software from a scratch. The last of them has no impact on stored data, whereas the rest of them can in some cases.
\paragraph{Add Property} Adding a property to database can change not only structure but stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{align*}
	\mathbf{addProperty} : Label \times Property \times State \rightarrow State \\
	\Psi(s, addProperty(l, p)) = application(s) \circ addProp(l,p) \\
	\Phi(s, addProperty(l, p)) = \begin{cases}
    		\inference{cardinality(p) \leq 1}{database(s) \bullet addColumn(l, \rho(p,d), d)} \\ \\
	    \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
   database(s) \bullet addTable(\rho(p,d), d) \\ \bullet \forall r \in row(owningClass(p)) :\\ insertRow(row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))])) 
    \end{gathered} } 
   \end{cases}\\
	s \lhd State, l \in Label, p \in Properties
\end{align*}

\paragraph{Add association} Adding an association into a database is complicated because of already existing instances. When an association between classes is added the instances of both classes has to be connected on the database level via foreign key or connection table. The information about pairing has to be provided by the user in form of set of instances pairs, because the model does not contain enough information. 
\begin{align*}
\mathbf{addAsociation} : Label \times Association \times (Row \times Row) \times State \rightarrow State \\
\Psi(s, addAsociation(l, a, m_{l, reference(a)})) = application(s) \circ addAssoc(l, a) \\
\Phi(s, addAsociation(l, a, m_{l, reference(a)})) = \begin{cases}
	\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = l\\ 		\wedge name(t_t) = reference(a) \\ \wedge \forall f \in foreignKeys(t_t) \cup columns(t_t) \\: name(f) \neq name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
		database(s) \\ \bullet addForeignKey(name(t_s), \rho(a,d), d) \\ \bullet \forall r \in m_{l, reference(a)} : \\ insertRow(row(name(t_s), [name(a), \\ value(primaryKey(first(r)))]), d)
	\end{gathered}}
 \\ \\
	\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = l \\ \wedge name(t_t) = reference(a) \\ \wedge startCardinality(a) > 1}{\begin{gathered} 
	database(s) \bullet addTable(\rho(a,d),d) \\ \bullet \forall r \in m_{l, reference(a)} :\\ insertRow(\\ row(name(owningClass(a)) + name(a), \\ [name(t_s), value(primaryKey(first(r)))]\\ [name(t_t), value(primaryKey(second(r)))]))
	\end{gathered}} 
\end{cases}\\
\end{align*}

@@

\paragraph{Remove property}
\begin{align*}
	\mathbf{RemoveProperty} : Label \times Label \times State \rightarrow State \\
	\Psi RemoveProperty(l, l_p, s) =  applications(s) \circ  remProp(l, l_p) \\
	\Phi RemoveProperty(l, l_p, s) = \begin{cases}
		 \inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) \leq 1}{database(s) \bullet dropColumn(l, l_p)}  \\\\
		 \inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists p \in properties(c)  : name(p) = l \\ \wedge cardinality(p)>1}{database(s) \bullet dropTable(name(owningClass(p)+l)}
 	\end{cases}\\
 	l, l_p \in Label, s \lhd State
\end{align*}

\paragraph{Remove Property and its Data} The $RemovePropertyAndData$ transformation helps to remove a property together with data stored in the database. The evolution can cause data and information lost, to prevent this the transformation $RemovePropertyAndData$ should be used carefully.
\begin{align*}
	\mathbf{RemovePropertyAndData} : Label \times Label* \times State \rightarrow State \\
	\Phi(RemovePropertyAndData(L_c, ps), s) = application(s)\\ \circ remProperty(l_s, ps) \\
	\Psi(RemovePropertyAndData(L_c, ps), s) = \begin{cases}
		%\inference{ps = \varnothing}{d}\\
		%\inference{|ps| > 1}{\begin{gathered} database(s) \\ \bullet RemovePropertyAndData(L_c, head(ps)) \\ \bullet RemovePropertyAndData(L_c, tail(ps)) 	\end{gathered}}\\\\
 		\inference{\exists c \in classes(application(s)) : name(c) = l\\ 	\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) \leq 1}{database(s) \bullet DropColumnAndData(l_c, ps)}\\\\
 	\inference{\exists c \in classes(application(s)) : name(c) = l\\ 	\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) > 1}{database(s) \bullet RemoveTableAndData(l_c, ps)}
 \end{cases}
	\\
	l_c \in Label, ps \lhd Label*, s \lhd State
\end{align*}

\paragraph{Remove association} 
\begin{align*}
    \mathbf{RemoveAssociation} : Label \times Labels \times State \rightarrow State \\ 
    \Psi RemoveAssociation(l, l_a, s) = application(s) \circ remAss(l, l_a) \\
    \Phi RemoveAssociation(l, l_a, s) = \begin{cases}
     	\inference{\exists c \in classes(application(s)) : name(c) = l \\
     	\wedge \exists a \in associaions(c) : name(a) = l_a  \\
     	\wedge startCardinality(a) \leq 1 \\}{database(s) \bullet dropForeignKey(l_a, reference(a))} \\\\
     	\inference{\exists c \in classes(application(s)) : name(c) = l \\
     	\wedge \exists a \in associaions(c) : name(a) = l_a  \\
     	\wedge startCardinality(a) > 1}{database(s) \bullet dropTable(name(owningClass(a))+l_a)} 
     \end{cases}\\
    l_c, l_a \in Labels, s \lhd State
\end{align*}
